---
title: "Lake Shore Drive Corridor Crash Analysis"
output:
  html_document:
    output_dir: docs
    output_file: dlsd_crashes.html
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Data Loading and Processing

First, we'll load our dependencies and process the crash data:

```{r load-data, message=FALSE, warning=FALSE}
# Source utility functions (which loads libraries)
source("R/utils.R")

# Read and process crash data
crashes <- load_crash_data("data/raw/crashes/traffic_crashes.csv")
crashes_sf <- convert_to_sf(crashes)

# Read and process fatality data
crash_fatalities <- load_fatality_data("data/raw/crashes/crash_fatalities.csv")
crash_fatalities_sf <- convert_to_sf(crash_fatalities, "Longitude", "Latitude")

# Combine crashes and fatalities
combined_crashes_sf <- dplyr::bind_rows(crashes_sf, crash_fatalities_sf)
```

## Lake Shore Drive Processing

Next, we'll process the Lake Shore Drive information:

```{r process-lsd, message=FALSE, warning=FALSE}
library(sf)
library(dplyr)
library(osmdata)
library(stringr)

# 1) Define a bounding box for Chicago and the buffer distance (150 ft ~ 45.72 m)
CHICAGO_BBOX <- c(-87.9401, 41.6445, -87.5239, 42.0230)
BUFFER_DISTANCE_M <- 45.72

# 2) Query OSM for the main US 41 portion of Lake Shore Drive
lsd_us41_query <- opq(bbox = CHICAGO_BBOX) %>%
  add_osm_feature(key = 'highway') %>%
  add_osm_feature(key = 'ref', value = 'US 41', value_exact = FALSE) %>%
  add_osm_feature(key = 'name', value = 'Lake Shore', value_exact = FALSE) %>%
  osmdata_sf()
lsd_us41_lines <- lsd_us41_query$osm_lines

# 3) Query OSM for the northern extension by name
lsd_north_query <- opq(bbox = CHICAGO_BBOX) %>%
  add_osm_feature(key = 'highway') %>%
  add_osm_feature(key = 'name', value = 'North Jean Baptiste Point DuSable Lake Shore Drive', value_exact = FALSE) %>%
  osmdata_sf()
lsd_north_lines <- lsd_north_query$osm_lines

# 4) Extract just the geometry from each to avoid mismatched columns
us41_geom  <- sf::st_geometry(lsd_us41_lines)
north_geom <- sf::st_geometry(lsd_north_lines)

# 5) Combine the geometry into a single sf object
combined_geom <- c(us41_geom, north_geom)
lake_shore_streets <- sf::st_sf(
  name = rep("Lake Shore Drive", length(combined_geom)),
  geometry = combined_geom,
  crs = sf::st_crs(lsd_us41_lines)
)

# 6) Transform to EPSG 3857 for accurate distance calculations
lake_shore_streets <- sf::st_transform(lake_shore_streets, 3857)
combined_crashes_sf <- sf::st_transform(combined_crashes_sf, 3857)
```

## Crash Analysis

Process crashes within the Lake Shore Drive buffer:

```{r analyze-crashes, message=FALSE, warning=FALSE}
# 1) Create buffer around Lake Shore Drive and identify intersecting crashes
buffer <- st_buffer(lake_shore_streets, dist = BUFFER_DISTANCE_M)
intersects <- st_intersects(combined_crashes_sf, buffer, sparse = FALSE)

# 2) Filter crashes within buffer
lakeshore_crashes <- combined_crashes_sf[apply(intersects, 1, any), ] %>%
  mutate(
    crash_count = 1,
    fatality_count = ifelse(!is.na(FATALITY_VICTIM) & FATALITY_VICTIM != "", 1, 0),
    total_injuries = INJURIES_TOTAL,
    incap_injuries = INJURIES_INCAPACITATING
  )

# 3) Create fatality subset
lakeshore_fatalities <- lakeshore_crashes %>%
  filter(fatality_count > 0)
```

## Study Area Definition

This analysis examines crashes within 150 feet (approximately 30.48
meters) of Lake Shore Drive. Here's a visualization of how we define the
study area:

```{r buffer-visualization, fig.width=10, fig.height=6, message=FALSE, warning=FALSE}
# Transform to WGS84 for visualization
lake_shore_streets_4326 <- st_transform(lake_shore_streets, 4326)
buffer_4326 <- st_transform(buffer, 4326)

# Create a zoomed-in example of the buffer
buffer_example <- leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  # Add Lake Shore Drive
  addPolylines(
    data = lake_shore_streets_4326,
    color = "#4B0082",  # Indigo
    weight = 3,
    label = "Lake Shore Drive"
  ) %>%
  # Add the buffer area
  addPolygons(
    data = buffer_4326,
    fillColor = "#9370DB",  # Medium purple
    fillOpacity = 0.3,
    color = "#4B0082",  # Indigo
    weight = 2,
    label = "150ft Buffer Zone"
  ) %>%
  # Set view to a section that shows the buffer clearly
  setView(lng = -87.6298, lat = 41.8781, zoom = 14)

# Display the map
buffer_example
```

The purple-shaded area represents our study area - a 150-foot buffer on
either side of Lake Shore Drive. All crash statistics in this analysis
are drawn from incidents that occurred within this buffer zone.

## Crash Impact Visualizations

### Fatalities Over Time

```{r fatalities-chart, fig.width=8, fig.height=5, message=FALSE, warning=FALSE}
# Prepare fatalities data
fatalities_by_year <- lakeshore_crashes %>%
  filter(fatality_count > 0) %>%
  st_drop_geometry() %>%
  mutate(year = year(CRASH_DATE)) %>%
  group_by(year) %>%
  summarize(total_fatalities = sum(fatality_count, na.rm = TRUE))

# Get min and max years for consistent x-axis
min_year <- min(fatalities_by_year$year)
max_year <- max(fatalities_by_year$year)

# Create the fatalities visualization
ggplot(fatalities_by_year, aes(x = year, y = total_fatalities)) +
  geom_col(fill = "#9370DB", width = 0.7) +
  geom_text(
    aes(label = total_fatalities),
    vjust = -0.5,
    size = 3.5,
    family = "Arial"
  ) +
  labs(
    x = "Year",
    y = NULL,
    title = "Traffic Fatalities Along Lake Shore Drive",
    subtitle = "Annual count of traffic-related fatalities",
    caption = "Data: Chicago City Data Portal \"Traffic Crashes - Vision Zero Chicago Traffic Fatalities\" Retrieved Feb. 6, 2025
Includes crashes reported within 150ft of DuSable Lake Shore Drive"
  ) +
  scale_x_continuous(
    breaks = seq(min_year, max_year, by = 1)
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.15)),
    breaks = function(x) seq(0, ceiling(max(x)), by = 5)
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(face = "bold", size = 12, margin = margin(b = 10)),
    plot.subtitle = element_text(size = 10, color = "#666666"),
    plot.background = element_rect(fill = "#FFFCF7"),
    panel.background = element_rect(fill = "#FFFCF7"),
    panel.grid.major.y = element_line(color = "#f0f0f0", linewidth = 0.3),
    text = element_text(family = "Arial"),
    axis.line = element_line(color = "#666666"),
    axis.ticks = element_line(color = "#666666"),
    axis.title = element_text(color = "#666666"),
    axis.text = element_text(color = "#666666"),
    plot.caption = element_text(
      color = "#666666",
      size = 8,
      margin = margin(t = 10),
      hjust = 0
    ),
    plot.margin = margin(t = 15, r = 20, b = 15, l = 15, unit = "pt")
  )
```

### Total Injuries Over Time

```{r injuries-chart, fig.width=8, fig.height=5, message=FALSE, warning=FALSE}
# Prepare injuries data
injuries_by_year <- lakeshore_crashes %>%
  st_drop_geometry() %>%
  mutate(
    year = year(CRASH_DATE),
    total_injuries = INJURIES_TOTAL
  ) %>%
  group_by(year) %>%
  summarize(total_injuries = sum(total_injuries, na.rm = TRUE))

# Create the injuries visualization
ggplot(injuries_by_year, aes(x = year, y = total_injuries)) +
  geom_col(fill = "#9370DB", width = 0.7) +
  geom_text(
    aes(label = total_injuries),
    vjust = -0.5,
    size = 3.5,
    family = "Arial"
  ) +
  labs(
    x = "Year",
    y = NULL,
    title = "Injuries From Crashes Along DuSable Lake Shore Drive",
    subtitle = "Total persons sustaining fatal, incapacitating, non-incapacitating, and possible injuries",
    caption = "Data: Chicago City Data Portal \"Traffic Crashes\" Retrieved Feb. 6, 2025
Includes crashes reported within 150ft of DuSable Lake Shore Drive"
  ) +
  scale_x_continuous(
    breaks = seq(min_year, max_year, by = 1)
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.15)),
    breaks = function(x) seq(0, ceiling(max(x)), by = 100)
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(face = "bold", size = 12, margin = margin(b = 10)),
    plot.subtitle = element_text(size = 10, color = "#666666"),
    plot.background = element_rect(fill = "#FFFCF7"),
    panel.background = element_rect(fill = "#FFFCF7"),
    panel.grid.major.y = element_line(color = "#f0f0f0", linewidth = 0.3),
    text = element_text(family = "Arial"),
    axis.line = element_line(color = "#666666"),
    axis.ticks = element_line(color = "#666666"),
    axis.title = element_text(color = "#666666"),
    axis.text = element_text(color = "#666666"),
    plot.caption = element_text(
      color = "#666666",
      size = 8,
      margin = margin(t = 10),
      hjust = 0
    ),
    plot.margin = margin(t = 15, r = 20, b = 15, l = 15, unit = "pt")
  )
```

### Incapacitating Injuries Over Time

```{r incap-injuries-chart, fig.width=8, fig.height=5, message=FALSE, warning=FALSE}
# Prepare incapacitating injuries data
incap_injuries_by_year <- lakeshore_crashes %>%
  st_drop_geometry() %>%
  mutate(
    year = year(CRASH_DATE),
    incap_injuries = INJURIES_INCAPACITATING
  ) %>%
  group_by(year) %>%
  summarize(total_incap_injuries = sum(incap_injuries, na.rm = TRUE))

# Create the incapacitating injuries visualization
ggplot(incap_injuries_by_year, aes(x = year, y = total_incap_injuries)) +
  geom_col(fill = "#9370DB", width = 0.7) +
  geom_text(
    aes(label = total_incap_injuries),
    vjust = -0.5,
    size = 3.5,
    family = "Arial"
  ) +
  labs(
    x = "Year",
    y = NULL,
    title = "Incapacitating Injuries Along DuSable Lake Shore Drive",
    subtitle = "Total persons sustaining incapacitating/serious injuries in a crash",
    caption = "Data: Chicago City Data Portal \"Traffic Crashes\" Retrieved Feb. 6, 2025
Includes crashes reported within 150ft of DuSable Lake Shore Drive"
  ) +
  scale_x_continuous(
    breaks = seq(min_year, max_year, by = 1)
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.15)),
    breaks = function(x) seq(0, ceiling(max(x)), by = 10)
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(face = "bold", size = 12, margin = margin(b = 10)),
    plot.subtitle = element_text(size = 10, color = "#666666"),
    plot.background = element_rect(fill = "#FFFCF7"),
    panel.background = element_rect(fill = "#FFFCF7"),
    panel.grid.major.y = element_line(color = "#f0f0f0", linewidth = 0.3),
    text = element_text(family = "Arial"),
    axis.line = element_line(color = "#666666"),
    axis.ticks = element_line(color = "#666666"),
    axis.title = element_text(color = "#666666"),
    axis.text = element_text(color = "#666666"),
    plot.caption = element_text(
      color = "#666666",
      size = 8,
      margin = margin(t = 10),
      hjust = 0
    ),
    plot.margin = margin(t = 15, r = 20, b = 15, l = 15, unit = "pt")
  )
```

## Summary Statistics

```{r crash-stats}
# Calculate Lake Shore Drive statistics
lakeshore_total_crashes <- nrow(lakeshore_crashes)
lakeshore_total_injuries <- sum(lakeshore_crashes$INJURIES_TOTAL, na.rm = TRUE)
lakeshore_total_incap_injuries <- sum(lakeshore_crashes$INJURIES_INCAPACITATING, na.rm = TRUE)
lakeshore_total_fatalities <- sum(lakeshore_crashes$fatality_count, na.rm = TRUE)

# Create summary statistics dataframe
crash_stats <- tibble(
  Metric = c("Total Crashes", "Total Injuries", "Incapacitating Injuries", "Fatalities"),
  `Lake Shore Drive Corridor` = c(lakeshore_total_crashes, lakeshore_total_injuries, lakeshore_total_incap_injuries, lakeshore_total_fatalities)
)

# Create formatted table
knitr::kable(crash_stats, format = "html") %>%
  kableExtra::kable_styling(
    bootstrap_options = "striped",
    full_width = FALSE,
    position = "left"
  ) %>%
  kableExtra::row_spec(0, bold = TRUE, color = "white", background = "#4B0082") %>%
  kableExtra::column_spec(1, bold = TRUE)
```

## Interactive Map Visualizations

### Map of Crashes with Injury

```{r injury-map}
# Transform to WGS84 for mapping
lakeshore_crashes_4326 <- st_transform(lakeshore_crashes, 4326)
lake_shore_streets_4326 <- st_transform(lake_shore_streets, 4326)

# Create binary indicators for crashes with injuries
lakeshore_crashes_4326 <- lakeshore_crashes_4326 %>%
  mutate(
    has_injuries = ifelse(INJURIES_TOTAL > 0, 1, 0),
    has_incap_injuries = ifelse(INJURIES_INCAPACITATING > 0, 1, 0)
  )

# Create injuries heatmap
injuries_map <- leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  # Add Lake Shore Drive
  addPolylines(
    data = lake_shore_streets_4326,
    color = "#666666",
    weight = 3,
    label = "Lake Shore Drive"
  ) %>%
  # Add injuries heatmap
  addHeatmap(
    data = lakeshore_crashes_4326 %>% filter(has_injuries == 1),
    lng = ~st_coordinates(geometry)[, 1],
    lat = ~st_coordinates(geometry)[, 2],
    intensity = ~1,  # Each crash with injuries counts as 1
    blur = 15,
    max = 0.1,
    radius = 12
  ) %>%
  setView(lng = -87.6298, lat = 41.8781, zoom = 12)

# Display the map
injuries_map
```

### Map of Crashes with Incapacitating Injury

```{r incap-injury-map}
# Create incapacitating injuries heatmap
incap_injuries_map <- leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  # Add Lake Shore Drive
  addPolylines(
    data = lake_shore_streets_4326,
    color = "#666666",
    weight = 3,
    label = "Lake Shore Drive"
  ) %>%
  # Add incapacitating injuries heatmap
  addHeatmap(
    data = lakeshore_crashes_4326 %>% filter(has_incap_injuries == 1),
    lng = ~st_coordinates(geometry)[, 1],
    lat = ~st_coordinates(geometry)[, 2],
    intensity = ~1,  # Each crash with incapacitating injuries counts as 1
    blur = 15,
    max = 0.1,
    radius = 12
  ) %>%
  setView(lng = -87.6298, lat = 41.8781, zoom = 12)

# Display the map
incap_injuries_map
```

### Map of Crashes with Fatalities

```{r fatality-map}
# Transform fatalities to WGS84 for mapping
lakeshore_fatalities_4326 <- st_transform(lakeshore_fatalities, 4326)

# Create fatalities map with clusters
fatalities_map <- leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  # Add Lake Shore Drive
  addPolylines(
    data = lake_shore_streets_4326,
    color = "#666666",
    weight = 3,
    label = "Lake Shore Drive"
  ) %>%
  # Add clustered fatality circles
  addCircleMarkers(
    data = lakeshore_fatalities_4326,
    lng = ~st_coordinates(geometry)[, 1],
    lat = ~st_coordinates(geometry)[, 2],
    radius = 6,
    color = "red",
    fillColor = "red",
    fillOpacity = 0.6,
    stroke = TRUE,
    weight = 1,
    popup = ~paste(
      "Date:", CRASH_DATE, "<br>",
      "Fatalities:", fatality_count, "<br>",
      "Type:", FIRST_CRASH_TYPE
    ),
    clusterOptions = markerClusterOptions(
      maxClusterRadius = 50 * 0.0003,
      spiderfyOnMaxZoom = TRUE,
      iconCreateFunction = JS("
        function(cluster) {
          var count = cluster.getChildCount();
          var size = Math.min(50, 20 + Math.sqrt(count) * 8);
          return new L.DivIcon({
            html: '<div style=\"background-color: rgba(255, 0, 0, 0.6); width: ' + size + 'px; height: ' + size + 'px; margin-left: -' + (size/2) + 'px; margin-top: -' + (size/2) + 'px; border-radius: 50%; display: flex; align-items: center; justify-content: center; color: white; font-weight: bold; border: 2px solid rgba(255, 0, 0, 0.8);\">' + count + '</div>',
            className: 'marker-cluster-custom',
            iconSize: L.point(size, size)
          });
        }
      ")
    )
  ) %>%
  setView(lng = -87.6298, lat = 41.8781, zoom = 12)

# Display the map
fatalities_map

```
