---
title: "Lake Shore Drive Corridor Crash Analysis"
output:
  html_document:
    output_dir: docs
    output_file: dlsd_crashes.html
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE, results='hide'}
knitr::opts_chunk$set(echo = FALSE)
```

```{r load-data, message=FALSE, warning=FALSE, results='hide'}
# Source utility functions (which loads libraries)
source("R/utils.R")
library(sf)
library(dplyr)
library(osmdata)
library(stringr)

# Read and process crash data
crashes <- load_crash_data("data/raw/crashes/traffic_crashes.csv")
crashes_sf <- convert_to_sf(crashes)

# Read and process fatality data
crash_fatalities <- load_fatality_data("data/raw/crashes/crash_fatalities.csv")
crash_fatalities_sf <- convert_to_sf(crash_fatalities, "Longitude", "Latitude")

# Combine crashes and fatalities
combined_crashes_sf <- dplyr::bind_rows(crashes_sf, crash_fatalities_sf)


# 1) Define a bounding box for Chicago and the buffer distance (150 ft ~ 45.72 m)
CHICAGO_BBOX <- c(-87.9401, 41.6445, -87.5239, 42.0230)
BUFFER_DISTANCE_M <- 45.72

# 2) Query OSM for the main US 41 portion of Lake Shore Drive
lsd_us41_query <- opq(bbox = CHICAGO_BBOX) %>%
  add_osm_feature(key = 'highway') %>%
  add_osm_feature(key = 'ref', value = 'US 41', value_exact = FALSE) %>%
  add_osm_feature(key = 'name', value = 'Lake Shore', value_exact = FALSE) %>%
  osmdata_sf()
lsd_us41_lines <- lsd_us41_query$osm_lines

# 3) Query OSM for the northern extension by name
lsd_north_query <- opq(bbox = CHICAGO_BBOX) %>%
  add_osm_feature(key = 'highway') %>%
  add_osm_feature(key = 'name', value = 'North Jean Baptiste Point DuSable Lake Shore Drive', value_exact = FALSE) %>%
  osmdata_sf()
lsd_north_lines <- lsd_north_query$osm_lines

# 4) Extract just the geometry from each to avoid mismatched columns
us41_geom  <- sf::st_geometry(lsd_us41_lines)
north_geom <- sf::st_geometry(lsd_north_lines)

# 5) Combine the geometry into a single sf object
combined_geom <- c(us41_geom, north_geom)
lake_shore_streets <- sf::st_sf(
  name = rep("Lake Shore Drive", length(combined_geom)),
  geometry = combined_geom,
  crs = sf::st_crs(lsd_us41_lines)
)

# 6) Transform to EPSG 3857 for accurate distance calculations
lake_shore_streets <- sf::st_transform(lake_shore_streets, 3857)
combined_crashes_sf <- sf::st_transform(combined_crashes_sf, 3857)

# 1) Create buffer around Lake Shore Drive and identify intersecting crashes
buffer <- st_buffer(lake_shore_streets, dist = BUFFER_DISTANCE_M)
intersects <- st_intersects(combined_crashes_sf, buffer, sparse = FALSE)

# 2) Filter crashes within buffer
lakeshore_crashes <- combined_crashes_sf[apply(intersects, 1, any), ] %>%
  mutate(
    crash_count = 1,
    fatality_count = ifelse(!is.na(FATALITY_VICTIM) & FATALITY_VICTIM != "", 1, 0),
    total_injuries = INJURIES_TOTAL,
    incap_injuries = INJURIES_INCAPACITATING
  )

# 3) Create fatality subset
lakeshore_fatalities <- lakeshore_crashes %>%
  filter(fatality_count > 0)
```

## Study Area Definition

This analysis examines crashes within 150 feet of Lake Shore Drive.
Here's a visualization of how we define the study area:

```{r buffer-visualization, fig.width=10, fig.height=6, message=FALSE, warning=FALSE}
# Transform to WGS84 for visualization
lake_shore_streets_4326 <- st_transform(lake_shore_streets, 4326)
buffer_4326 <- st_transform(buffer, 4326)

# Create a zoomed-in example of the buffer
buffer_example <- leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  # Add Lake Shore Drive
  addPolylines(
    data = lake_shore_streets_4326,
    color = "#4B0082",  # Indigo
    weight = 3,
    label = "Lake Shore Drive"
  ) %>%
  # Add the buffer area
  addPolygons(
    data = buffer_4326,
    fillColor = "#9370DB",  # Medium purple
    fillOpacity = 0.3,
    color = "#4B0082",  # Indigo
    weight = 2,
    label = "150ft Buffer Zone"
  ) %>%
  # Set view to a section that shows the buffer clearly
  setView(lng = -87.6298, lat = 41.8781, zoom = 14)

# Display the map
buffer_example
```

## Summary Statistics

Lake Shore Drive represents less than 0.5% of center-line length of all
roadways in Chicago, but it accounts for \~7% of all traffic fatalities.
It is a disproportionately fatal corridor.

```{r crash-stats, message=FALSE,}
# Load required libraries
library(dplyr)
library(knitr)
library(kableExtra)
library(sf)

# Calculate Lake Shore Drive statistics
# (lakeshore_crashes already has fatality_count from the previous step)
lakeshore_total_crashes <- nrow(lakeshore_crashes)
lakeshore_total_injuries <- sum(lakeshore_crashes$INJURIES_TOTAL, na.rm = TRUE)
lakeshore_total_incap_injuries <- sum(lakeshore_crashes$INJURIES_INCAPACITATING, na.rm = TRUE)
lakeshore_total_fatalities <- sum(lakeshore_crashes$fatality_count, na.rm = TRUE)

# Calculate citywide statistics
# First create fatality_count for citywide data the same way
crashes_with_fatalities <- combined_crashes_sf %>%
  mutate(
    fatality_count = ifelse(!is.na(FATALITY_VICTIM) & FATALITY_VICTIM != "", 1, 0)
  )

citywide_total_crashes <- nrow(crashes_with_fatalities)
citywide_total_injuries <- sum(crashes_with_fatalities$INJURIES_TOTAL, na.rm = TRUE)
citywide_total_incap_injuries <- sum(crashes_with_fatalities$INJURIES_INCAPACITATING, na.rm = TRUE)
citywide_total_fatalities <- sum(crashes_with_fatalities$fatality_count, na.rm = TRUE)

# Calculate percentages
pct_crashes <- round(lakeshore_total_crashes / citywide_total_crashes * 100, 1)
pct_injuries <- round(lakeshore_total_injuries / citywide_total_injuries * 100, 1)
pct_incap_injuries <- round(lakeshore_total_incap_injuries / citywide_total_incap_injuries * 100, 1)
pct_fatalities <- round(lakeshore_total_fatalities / citywide_total_fatalities * 100, 1)

# Create expanded summary statistics dataframe
crash_stats_expanded <- tibble(
  Metric = c("Total Crashes", "Total Injuries", "Incapacitating Injuries", "Fatalities"),
  `DLSD Corridor` = c(
    formatC(lakeshore_total_crashes, big.mark = ",", format = "f", digits = 0),
    formatC(lakeshore_total_injuries, big.mark = ",", format = "f", digits = 0),
    formatC(lakeshore_total_incap_injuries, big.mark = ",", format = "f", digits = 0),
    formatC(lakeshore_total_fatalities, big.mark = ",", format = "f", digits = 0)
  ),
  `Citywide Total` = c(
    formatC(citywide_total_crashes, big.mark = ",", format = "f", digits = 0),
    formatC(citywide_total_injuries, big.mark = ",", format = "f", digits = 0),
    formatC(citywide_total_incap_injuries, big.mark = ",", format = "f", digits = 0),
    formatC(citywide_total_fatalities, big.mark = ",", format = "f", digits = 0)
  ),
  `DLSD % of Citywide Total` = c(
    paste0(pct_crashes, "%"),
    paste0(pct_injuries, "%"),
    paste0(pct_incap_injuries, "%"),
    paste0(pct_fatalities, "%")
  )
)

# Create formatted table
crash_stats_table <- knitr::kable(crash_stats_expanded, 
                                  format = "html",
                                  caption = "<strong style='color: #333333;'>DLSD Summary 2019 - 2024</strong>") %>%
  kableExtra::kable_styling(
    bootstrap_options = "striped",
    full_width = FALSE,
    position = "left"
  ) %>%
  kableExtra::row_spec(0, bold = TRUE, color = "white", background = "#4B0082") %>%
  kableExtra::column_spec(1, bold = TRUE)

# Display the table
crash_stats_table
```

## Annual Trends

### Traffic Fatalities

```{r fatalities-chart, fig.width=8, fig.height=5, message=FALSE, warning=FALSE}
# Prepare fatalities data
fatalities_by_year <- lakeshore_crashes %>%
  filter(fatality_count > 0) %>%
  st_drop_geometry() %>%
  mutate(year = year(CRASH_DATE)) %>%
  group_by(year) %>%
  summarize(total_fatalities = sum(fatality_count, na.rm = TRUE))

# Get min and max years for consistent x-axis
min_year <- min(fatalities_by_year$year)
max_year <- max(fatalities_by_year$year)

# Create the fatalities visualization
ggplot(fatalities_by_year, aes(x = year, y = total_fatalities)) +
  geom_col(fill = "#9370DB", width = 0.7) +
  geom_text(
    aes(label = total_fatalities),
    vjust = -0.5,
    size = 3.5,
    family = "Arial"
  ) +
  labs(
    x = "Year",
    y = NULL,
    title = "Traffic Fatalities Along Lake Shore Drive",
    subtitle = "Annual count of traffic-related fatalities",
    caption = "Data: Chicago City Data Portal \"Traffic Crashes - Vision Zero Chicago Traffic Fatalities\" Retrieved Feb. 6, 2025
Includes crashes reported within 150ft of DuSable Lake Shore Drive"
  ) +
  scale_x_continuous(
    breaks = seq(min_year, max_year, by = 1)
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.15)),
    breaks = function(x) seq(0, ceiling(max(x)), by = 5)
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(face = "bold", size = 12, margin = margin(b = 10)),
    plot.subtitle = element_text(size = 10, color = "#666666"),
    plot.background = element_rect(fill = "#FFFCF7"),
    panel.background = element_rect(fill = "#FFFCF7"),
    panel.grid.major.y = element_line(color = "#f0f0f0", linewidth = 0.3),
    text = element_text(family = "Arial"),
    axis.line = element_line(color = "#666666"),
    axis.ticks = element_line(color = "#666666"),
    axis.title = element_text(color = "#666666"),
    axis.text = element_text(color = "#666666"),
    plot.caption = element_text(
      color = "#666666",
      size = 8,
      margin = margin(t = 10),
      hjust = 0
    ),
    plot.margin = margin(t = 15, r = 20, b = 15, l = 15, unit = "pt")
  )
```

### Total Injuries

```{r injuries-chart, fig.width=8, fig.height=5, message=FALSE, warning=FALSE}
# Prepare injuries data
injuries_by_year <- lakeshore_crashes %>%
  st_drop_geometry() %>%
  mutate(
    year = year(CRASH_DATE),
    total_injuries = INJURIES_TOTAL
  ) %>%
  group_by(year) %>%
  summarize(total_injuries = sum(total_injuries, na.rm = TRUE))

# Create the injuries visualization
ggplot(injuries_by_year, aes(x = year, y = total_injuries)) +
  geom_col(fill = "#9370DB", width = 0.7) +
  geom_text(
    aes(label = total_injuries),
    vjust = -0.5,
    size = 3.5,
    family = "Arial"
  ) +
  labs(
    x = "Year",
    y = NULL,
    title = "Injuries From Crashes Along DuSable Lake Shore Drive",
    subtitle = "Total persons sustaining fatal, incapacitating, non-incapacitating, and possible injuries",
    caption = "Data: Chicago City Data Portal \"Traffic Crashes\" Retrieved Feb. 6, 2025
Includes crashes reported within 150ft of DuSable Lake Shore Drive"
  ) +
  scale_x_continuous(
    breaks = seq(min_year, max_year, by = 1)
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.15)),
    breaks = function(x) seq(0, ceiling(max(x)), by = 100)
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(face = "bold", size = 12, margin = margin(b = 10)),
    plot.subtitle = element_text(size = 10, color = "#666666"),
    plot.background = element_rect(fill = "#FFFCF7"),
    panel.background = element_rect(fill = "#FFFCF7"),
    panel.grid.major.y = element_line(color = "#f0f0f0", linewidth = 0.3),
    text = element_text(family = "Arial"),
    axis.line = element_line(color = "#666666"),
    axis.ticks = element_line(color = "#666666"),
    axis.title = element_text(color = "#666666"),
    axis.text = element_text(color = "#666666"),
    plot.caption = element_text(
      color = "#666666",
      size = 8,
      margin = margin(t = 10),
      hjust = 0
    ),
    plot.margin = margin(t = 15, r = 20, b = 15, l = 15, unit = "pt")
  )
```

### Incapacitating Injuries

```{r incap-injuries-chart, fig.width=8, fig.height=5, message=FALSE, warning=FALSE}
# Prepare incapacitating injuries data
incap_injuries_by_year <- lakeshore_crashes %>%
  st_drop_geometry() %>%
  mutate(
    year = year(CRASH_DATE),
    incap_injuries = INJURIES_INCAPACITATING
  ) %>%
  group_by(year) %>%
  summarize(total_incap_injuries = sum(incap_injuries, na.rm = TRUE))

# Create the incapacitating injuries visualization
ggplot(incap_injuries_by_year, aes(x = year, y = total_incap_injuries)) +
  geom_col(fill = "#9370DB", width = 0.7) +
  geom_text(
    aes(label = total_incap_injuries),
    vjust = -0.5,
    size = 3.5,
    family = "Arial"
  ) +
  labs(
    x = "Year",
    y = NULL,
    title = "Incapacitating Injuries Along DuSable Lake Shore Drive",
    subtitle = "Total persons sustaining incapacitating/serious injuries in a crash",
    caption = "Data: Chicago City Data Portal \"Traffic Crashes\" Retrieved Feb. 6, 2025
Includes crashes reported within 150ft of DuSable Lake Shore Drive"
  ) +
  scale_x_continuous(
    breaks = seq(min_year, max_year, by = 1)
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.15)),
    breaks = function(x) seq(0, ceiling(max(x)), by = 10)
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(face = "bold", size = 12, margin = margin(b = 10)),
    plot.subtitle = element_text(size = 10, color = "#666666"),
    plot.background = element_rect(fill = "#FFFCF7"),
    panel.background = element_rect(fill = "#FFFCF7"),
    panel.grid.major.y = element_line(color = "#f0f0f0", linewidth = 0.3),
    text = element_text(family = "Arial"),
    axis.line = element_line(color = "#666666"),
    axis.ticks = element_line(color = "#666666"),
    axis.title = element_text(color = "#666666"),
    axis.text = element_text(color = "#666666"),
    plot.caption = element_text(
      color = "#666666",
      size = 8,
      margin = margin(t = 10),
      hjust = 0
    ),
    plot.margin = margin(t = 15, r = 20, b = 15, l = 15, unit = "pt")
  )
```

```{r setup-senate, message=FALSE, warning=FALSE}
# Load senate district shapefile and filter for lakefront districts
senate_districts <- load_district_shapefile(
  "data/raw/shapefiles/senate/senate_districts.shp",
  new_name = "senate_district"
)
lakefront_districts <- senate_districts %>%
  filter(senate_district %in% c("6", "7", "13"))

# Transform lakefront districts to EPSG:3857 to match the crashes data
lakefront_districts_3857 <- st_transform(lakefront_districts, 3857)

# Now perform the spatial join
lakefront_crashes <- sf::st_join(lakeshore_crashes, lakefront_districts_3857, left = FALSE)

# Prepare yearly data by district
crashes_by_district <- lakefront_crashes %>%
  st_drop_geometry() %>%
  mutate(
    year = year(CRASH_DATE),
    crash_count = 1,
    injury_count = INJURIES_TOTAL,
    incap_injury_count = INJURIES_INCAPACITATING,
    fatality_count = ifelse(!is.na(FATALITY_VICTIM) & FATALITY_VICTIM != "", 1, 0)
  ) %>%
  group_by(year, senate_district) %>%
  summarize(
    total_crashes = sum(crash_count),
    total_injuries = sum(injury_count, na.rm = TRUE),
    total_incap_injuries = sum(incap_injury_count, na.rm = TRUE),
    total_fatalities = sum(fatality_count, na.rm = TRUE)
  ) %>%
  ungroup()

# Function to create consistent chart theme
dlsd_theme <- function() {
  theme_classic() +
    theme(
      plot.title = element_text(face = "bold", size = 10, margin = margin(b = 10)),
      plot.subtitle = element_text(size = 10, color = "#666666"),
      plot.background = element_rect(fill = "#FFFCF7"),
      panel.background = element_rect(fill = "#FFFCF7"),
      panel.grid.major.y = element_line(color = "#f0f0f0", linewidth = 0.3),
      text = element_text(family = "Arial"),
      axis.line = element_line(color = "#666666"),
      axis.ticks = element_line(color = "#666666"),
      axis.title = element_text(color = "#666666"),
      axis.text = element_text(color = "#666666"),
      plot.caption = element_text(
        color = "#666666",
        size = 9,
        margin = margin(t = 10),
        hjust = 0
      ),
      plot.margin = margin(t = 15, r = 20, b = 15, l = 15, unit = "pt"),
      legend.position = "bottom",
      legend.title = element_blank()
    )
}

# Define consistent colors for districts
district_colors <- c(
  "6" = "#9370DB",
  "7" = "#0173b2",
  "13" = "#de8f05"
)

```

### DuSable Lake Shore Drive Crashes by Senate District

```{r}
# Calculate metrics by district
district_metrics <- lakefront_crashes %>%
  st_drop_geometry() %>%
  group_by(senate_district) %>%
  summarize(
    # Crashes
    crashes = n(),
    injury_crashes = sum(INJURIES_TOTAL > 0, na.rm = TRUE),
    hit_run_crashes = sum(HIT_AND_RUN_I == "Y", na.rm = TRUE),
    
    # Injuries
    injury_count = sum(INJURIES_TOTAL, na.rm = TRUE),
    hit_run_injuries = sum(ifelse(HIT_AND_RUN_I == "Y", INJURIES_TOTAL, 0), na.rm = TRUE),
    incap_injuries = sum(INJURIES_INCAPACITATING, na.rm = TRUE),
    
    # Fatalities
    fatalities = sum(fatality_count, na.rm = TRUE)
  ) %>%
  ungroup()

# Create the data frame structure
summary_table <- data.frame(
  "Severity" = c(
    rep("Crashes", 3),
    rep("Injuries", 3),
    rep("Fatalities", 1)
  ),
  "Metric" = c(
    # Crashes section
    "Crashes",
    "Crashes with Injury",
    "Hit and Run Crashes",
    # Injuries section
    "Injury Count",
    "Hit and Run Injuries",
    "Incapacitating Injury Count",
    # Fatalities section
    "Fatalities"
  )
)

# Function to get metrics in the right order
get_metrics <- function(district_data) {
  c(
    district_data$crashes,
    district_data$injury_crashes,
    district_data$hit_run_crashes,
    district_data$injury_count,
    district_data$hit_run_injuries,
    district_data$incap_injuries,
    district_data$fatalities
  )
}

# Add district columns with actual data
district_data <- data.frame(
  "District 13" = get_metrics(filter(district_metrics, senate_district == "13")),
  "District 6" = get_metrics(filter(district_metrics, senate_district == "6")),
  "District 7" = get_metrics(filter(district_metrics, senate_district == "7")),
  check.names = FALSE
)

# Calculate totals
district_data$Total <- rowSums(district_data)

# Format numbers with commas
district_data[] <- lapply(district_data, function(x) formatC(x, big.mark = ",", format = "f", digits = 0))

# Combine all data
summary_table <- cbind(summary_table, district_data)

# Create the formatted table
crash_stats_table <- knitr::kable(
  summary_table,
  format = "html",
  caption = "<strong style='color: #333333;'>DuSable Lake Shore Drive Crash Report - Senate</strong>
             <br><em>Jan 2019 - Dec 2024</em>"
) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "bordered"),
    full_width = TRUE,
    position = "left"
  ) %>%
  kableExtra::row_spec(0, bold = TRUE, color = "white", background = "#4B0082") %>%
  kableExtra::column_spec(1:2, bold = TRUE) %>%
  kableExtra::collapse_rows(
    columns = 1,
    valign = "top"
  )

# Display the table
crash_stats_table
```

### Annual Crashes by Senate District

```{r crashes-by-district, fig.width=10, fig.height=6}
ggplot(crashes_by_district, 
       aes(x = year, y = total_crashes, fill = senate_district)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.7) +
  geom_text(
    aes(label = scales::comma(total_crashes)),  # Added thousands separator
    position = position_dodge(width = 0.7),
    vjust = -0.5,
    size = 3,
    family = "Arial"
  ) +
  scale_fill_manual(
    values = district_colors,
    breaks = c("13", "7", "6"),
    labels = c("District 13", "District 7", "District 6")
  ) +
  labs(
    x = "Year",
    y = NULL,
    title = "DuSable Lake Shore Drive Crashes by Senate District",
    subtitle = "Annual count of reported crashes in the E-Crash CPD database",
    caption = "Data: Chicago City Data Portal \"Traffic Crashes\" Retrieved Feb. 6, 2025
Study area: Includes crashes reported within 150ft of DuSable Lake Shore Drive"
  ) +
  scale_x_continuous(
    breaks = seq(min(crashes_by_district$year), max(crashes_by_district$year), by = 1)  # Show every year
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.15))
  ) +
  dlsd_theme() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),  # Explicitly set larger title size here
    plot.caption = element_text(
        color = "#666666",
        size = 10,
        margin = margin(t = 10),
        hjust = 0
      )
  )
```

### Injuries by Senate District

```{r injuries-by-district, fig.width=10, fig.height=6}
ggplot(crashes_by_district, 
       aes(x = year, y = total_injuries, fill = senate_district)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.7) +
  geom_text(
    aes(label = scales::comma(total_injuries)),
    position = position_dodge(width = 0.7),
    vjust = -0.5,
    size = 3,
    family = "Arial"
  ) +
  scale_fill_manual(
    values = district_colors,
    breaks = c("13", "7", "6"),
    labels = c("District 13", "District 7", "District 6")
  ) +
  labs(
    x = "Year",
    y = NULL,
    title = "DuSable Lake Shore Drive Injuries by Senate District",
    subtitle = "Total persons sustaining fatal, incapacitating, non-incapacitating, and possible injuries",
    caption = "Data: Chicago City Data Portal \"Traffic Crashes\" Retrieved Feb. 6, 2025
Study area: Includes crashes reported within 150ft of DuSable Lake Shore Drive
Description: Total persons sustaining fatal, incapacitating, non-incapacitating, and possible injuries as determined by the reporting officer"
  ) +
  scale_x_continuous(
    breaks = seq(min(crashes_by_district$year), max(crashes_by_district$year), by = 1)  # Show every year
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.15))
  ) + 
  dlsd_theme() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),  # Explicitly set larger title size here
    plot.subtitle = element_text(size = 12, color = "#666666"),
    plot.caption = element_text(
        color = "#666666",
        size = 10,
        margin = margin(t = 10),
        hjust = 0
      )
  )
```

### Incapacitating Injuries by Senate District

```{r incap-injuries-by-district, fig.width=10, fig.height=6}
ggplot(crashes_by_district, 
       aes(x = year, y = total_incap_injuries, fill = senate_district)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.7) +
  geom_text(
    aes(label = scales::comma(total_incap_injuries)),
    position = position_dodge(width = 0.7),
    vjust = -0.5,
    size = 3,
    family = "Arial"
  ) +
  scale_fill_manual(
    values = district_colors,
    breaks = c("13", "7", "6"),
    labels = c("District 13", "District 7", "District 6")
  ) +
  labs(
    x = "Year",
    y = NULL,
    title = "DuSable Lake Shore Drive Incapacitating Injuries by Senate District",
    subtitle = "Total persons sustaining incapacitating/serious injuries",
    caption = "Data: Chicago City Data Portal \"Traffic Crashes\" Retrieved Feb. 6, 2025
Study area: Includes crashes reported within 150ft of DuSable Lake Shore Drive
Description: Any injury other than fatal injury, which prevents the injured person from walking, driving, or normally continuing the activities they were 
capable of performing before the injury occurred. Includes severe lacerations, broken limbs, skull or chest injuries, and abdominal injuries."
  ) +
  scale_x_continuous(
    breaks = seq(min(crashes_by_district$year), max(crashes_by_district$year), by = 1)  # Show every year
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.15))
  ) +
  dlsd_theme() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),  # Explicitly set larger title size here
    plot.subtitle = element_text(size = 12, color = "#666666"),
    plot.caption = element_text(
        color = "#666666",
        size = 10,
        margin = margin(t = 10),
        hjust = 0
      )
  )

```

### Fatalities by Senate District

```{r fatalities-by-district, fig.width=10, fig.height=6}
ggplot(crashes_by_district, 
       aes(x = year, y = total_fatalities, fill = senate_district)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.7) +
  geom_text(
    aes(label = scales::comma(total_fatalities)),
    position = position_dodge(width = 0.7),
    vjust = -0.5,
    size = 3,
    family = "Arial"
  ) +
  scale_fill_manual(
    values = district_colors,
    breaks = c("13", "7", "6"),
    labels = c("District 13", "District 7", "District 6")
  ) +
  labs(
    x = "Year",
    y = NULL,
    title = "DuSable Lake Shore Drive Fatalities by Senate District",
    subtitle = "Annual count of reported traffic-related fatalities",
    caption = "Data: Chicago City Data Portal \"Traffic Crashes - Vision Zero Chicago Traffic Fatalities\" Retrieved Feb. 6, 2025
Study area: Includes crashes reported within 150ft of DuSable Lake Shore Drive"
  ) +
  scale_x_continuous(
    breaks = seq(min(crashes_by_district$year), max(crashes_by_district$year), by = 1)  # Show every year
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.15))
  ) +
  dlsd_theme() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),  # Explicitly set larger title size here
    plot.subtitle = element_text(size = 12, color = "#666666"),
    plot.caption = element_text(
        color = "#666666",
        size = 10,
        margin = margin(t = 10),
        hjust = 0
      )
  )
```

## Crash Clusters

### Where are crashes happening most often?

Hover over a tile to see the summary of crashes, injuries, and
fatalities.

```{r}
# Transform data back to WGS84 for mapping
lakefront_crashes_4326 <- st_transform(lakefront_crashes, 4326)
lakefront_districts_4326 <- st_transform(lakefront_districts, 4326)

# Create binary indicators for different crash types
lakefront_crashes_4326 <- lakefront_crashes_4326 %>%
  mutate(
    has_injuries = ifelse(INJURIES_TOTAL > 0, 1, 0),
    has_incap_injuries = ifelse(INJURIES_INCAPACITATING > 0, 1, 0),
    has_fatality = ifelse(!is.na(FATALITY_VICTIM) & FATALITY_VICTIM != "", 1, 0)
  )

# Transform to UTM Zone 16N for grid creation
bbox_utm <- st_transform(st_as_sfc(st_bbox(lakefront_crashes_4326)), 26916)

# Create hex grid in UTM with cell size in meters
hex_grid <- st_make_grid(bbox_utm,
                        cellsize = c(500, 500),  # 500 meter hexagons
                        square = FALSE) %>%
  st_sf() %>%
  st_transform(4326)  # Transform back to WGS84 for Leaflet

# Create the hex grid with additional statistics
hex_sf <- hex_grid %>%
  mutate(
    count = lengths(st_intersects(., lakefront_crashes_4326)),
    total_injuries = sapply(st_intersects(., lakefront_crashes_4326), function(idx) {
      sum(lakefront_crashes_4326$INJURIES_TOTAL[idx], na.rm = TRUE)
    }),
    injuries_incap = sapply(st_intersects(., lakefront_crashes_4326), function(idx) {
      sum(lakefront_crashes_4326$INJURIES_INCAPACITATING[idx], na.rm = TRUE)
    }),
    fatalities = sapply(st_intersects(., lakefront_crashes_4326), function(idx) {
      sum(lakefront_crashes_4326$has_fatality[idx], na.rm = TRUE)
    })
  ) %>%
  filter(count > 0)

# Create color palette
pal <- colorNumeric(
  palette = "YlOrRd",  # Yellow-Orange-Red color scheme
  domain = hex_sf$count
)

# Create the map
crash_map <- leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    data = lakefront_districts_4326,
    fillColor = "transparent",
    weight = 2,
    color = "#666666",
    dashArray = "3",
    label = ~paste("Senate District", senate_district)
  ) %>%
  addPolylines(
    data = lake_shore_streets_4326,
    color = "#666666",
    weight = 3,
    label = "Lake Shore Drive"
  ) %>%
  addPolygons(
    data = hex_sf,
    fillColor = ~pal(count),
    fillOpacity = 0.7,
    color = "#FFFFFF",
    weight = 1,
    label = ~sprintf(
      "%d crashes<br>%d injuries<br>%d incapacitating injuries<br>%d fatalities",
      count,
      total_injuries,
      injuries_incap,
      fatalities
    ) %>% lapply(htmltools::HTML),
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.9,
      bringToFront = TRUE
    )
  ) %>%
  addLegend(
    position = "bottomright",
    pal = pal,
    values = hex_sf$count,
    title = "Crash Count",
    opacity = 0.7
  ) %>%
  setView(lng = -87.6298, lat = 41.8781, zoom = 12)

# Display the map
crash_map
```

### Where are incapacitating injuries happening most often?

```{r}
# Transform to UTM Zone 16N for grid creation
bbox_utm <- st_transform(st_as_sfc(st_bbox(lakefront_crashes_4326)), 26916)

# Create hex grid in UTM with cell size in meters
hex_grid <- st_make_grid(bbox_utm, 
                        cellsize = c(450, 450),  # 200 meter hexagons
                        square = FALSE) %>%
  st_sf() %>%
  st_transform(4326)  # Transform back to WGS84 for Leaflet

# Calculate incapacitating injury counts per hexagon
hex_incap_counts <- sapply(st_intersects(hex_grid, lakefront_crashes_4326), function(idx) {
  sum(lakefront_crashes_4326$incap_injuries[idx], na.rm = TRUE)
})

# Create the hex grid with incapacitating injury statistics
hex_incap_sf <- hex_grid %>%
  mutate(
    incap_count = hex_incap_counts
  ) %>%
  filter(incap_count > 0)

# Create color palette for incapacitating injuries
pal_incap <- colorNumeric(
  palette = "YlOrRd",
  domain = hex_incap_sf$incap_count
)

# Create the incapacitating injuries map
incap_map <- leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    data = lakefront_districts_4326,
    fillColor = "transparent",
    weight = 2,
    color = "#666666",
    dashArray = "3",
    label = ~paste("Senate District", senate_district)
  ) %>%
  addPolylines(
    data = lake_shore_streets_4326,
    color = "#666666",
    weight = 3,
    label = "Lake Shore Drive"
  ) %>%
  addPolygons(
    data = hex_incap_sf,
    fillColor = ~pal_incap(incap_count),
    fillOpacity = 0.7,
    color = "#FFFFFF",
    weight = 1,
    label = ~sprintf(
      "%d incapacitating injuries",
      incap_count
    ) %>% lapply(htmltools::HTML),
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.9,
      bringToFront = TRUE
    )
  ) %>%
  addLegend(
    position = "bottomright",
    pal = pal_incap,
    values = hex_incap_sf$incap_count,
    title = "Incapacitating Injuries",
    opacity = 0.7
  ) %>%
  setView(lng = -87.6298, lat = 41.8781, zoom = 12)
incap_map
```

### Where are fatal crashes occurring?

```{r}
# Calculate fatality counts per hexagon
hex_fatal_counts <- sapply(st_intersects(hex_grid, lakefront_crashes_4326), function(idx) {
  sum(lakefront_crashes_4326$fatality_count[idx], na.rm = TRUE)
})

# Create the hex grid with fatality statistics
hex_fatal_sf <- hex_grid %>%
  mutate(
    fatal_count = hex_fatal_counts
  ) %>%
  filter(fatal_count > 0)

# Create color palette for fatalities
pal_fatal <- colorNumeric(
  palette = c("#fb6a4a","#ef3b2c","#cb181d","#a50f15","#67000d"),
  domain = hex_fatal_sf$fatal_count
)

# Create the fatalities map
fatal_map <- leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    data = lakefront_districts_4326,
    fillColor = "transparent",
    weight = 2,
    color = "#666666",
    dashArray = "3",
    label = ~paste("Senate District", senate_district)
  ) %>%
  addPolylines(
    data = lake_shore_streets_4326,
    color = "#666666",
    weight = 3,
    label = "Lake Shore Drive"
  ) %>%
  addPolygons(
    data = hex_fatal_sf,
    fillColor = ~pal_fatal(fatal_count),
    fillOpacity = 0.7,
    color = "#FFFFFF",
    weight = 1,
    label = ~sprintf(
      "%d fatalities",
      fatal_count
    ) %>% lapply(htmltools::HTML),
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.9,
      bringToFront = TRUE
    )
  ) %>%
  addLegend(
    position = "bottomright",
    pal = pal_fatal,
    values = hex_fatal_sf$fatal_count,
    title = "Fatalities",
    opacity = 0.7
  ) %>%
  setView(lng = -87.6298, lat = 41.8781, zoom = 12)
fatal_map
```
